---
title: "hw8-Generalized Linear models for longitudinal Data case study"
author: "Cary Ni"
output: 
  pdf_document :
    latex_engine: xelatex
---

```{r setup, include=FALSE}
library(tidyverse)
library(nlme)
library(lattice)
library(lme4)
library(gee)
knitr::opts_chunk$set(echo = TRUE)
```

```{r load dataset}
# load dataset and modify variable types
data_df = readxl::read_excel("./data/HW8-HEALTH.xlsx") %>% janitor::clean_names()
data_df[, c(3:5)]= lapply(data_df[, c(3:5)], as.factor)
summary(data_df)
```

```{r}
# data wrangling
# 1 == good
data_base = subset(data_df, time == 1) %>% 
  mutate(baseline = health) %>% 
  mutate(n_health = as.numeric(health == "Good"))

data_sub = data_base %>% select(id, baseline)

data_new = data_df %>% 
  left_join(data_sub, by = "id") %>% 
  filter(time != 1) %>% 
  mutate(n_health = as.numeric(health == "Good"))
 
```

## (a) 

```{r}
# fit cross-sectional model at randomization 
health_glm = glm(n_health ~ txt, data = data_base, family = "binomial")
summary(health_glm)
```

At the time of randomization, it can be seen that odds of being good in `health` is lower for the intervention group compared to control group with a factor of `r round(exp(-0.314), 3)`. However, the p value gives 0.486, meaning that this difference between two groups is not significant at 0.05 significance level. We can say that the baseline odds of being good in
`health` is not significantly different for intervention group and control group.

## (b)

```{r}
# fit gee after randomization
health_gee = gee(n_health ~ baseline + txt + time + agegroup, data = data_new, 
                 family = "binomial", id = id, corstr = "unstructured", scale.fix = FALSE)
summary(health_gee)
```

From the GEE model with unstructured correlation, it can be seen from coefficients that the baseline of being poor decrease the odds of being good in level of health to a factor of `r round(exp(-1.8144), 3)` compared to being good in baseline while holding other covariates fixed. This term is also shown to be significant as its p value is less than 0.05 at the given significance level. 
Holding other covariates fixed, the intervention group is shown to be have a `r round(exp(2.1), 3)` times odds of reporting good in level of health compared to control group with a p value less than 0.05 also suggests its significance. 
Holding other covariates fixed, with every unit increase in time (3 months), the odds of reporting good increase to a factor of `r round(exp(0.153), 3)` though this term is shown to be insignificant with a p value larger than 0.05. 
Holding other covariates fixed, compared to age group (15-24), age group (25-34) has `r round(exp(1.35), 3)` times the odds of reporting good with significance in p value less than 0.05, whereas age group (35+) is not significantly different from the group (15-24) with a p value larger than 0.05 even though it has `r round(exp(1.412), 3)` the odds
of reporting good compared to group (15-24). 

## (c)

```{r}
# Fit the model with random intercept
health_glmm = glmer(n_health ~ baseline + txt + time + agegroup + (1 | id), 
                    data = data_new, 
                    family = "binomial")
summary(health_glmm)
```

From the general linear mixed effects model, it can be seen from coefficients that the baseline of being poor decrease the odds of being good in level of health to a factor of `r round(exp(-2.78), 3)` compared to being good in baseline while holding other covariates fixed. This term is also shown to be significant as its p value is less than 0.05 at the given significance level. 
Holding other covariates fixed, the intervention group is shown to be have a `r round(exp(3.42), 3)` times odds of reporting good in level of health compared to control group with a p value less than 0.05 also suggests its significance. 
Holding other covariates fixed, with every unit increase in time (3 months), the odds of reporting good increase to a factor of `r round(exp(0.2022), 3)` though this term is shown to be insignificant with a p value larger than 0.05. 
Holding other covariates fixed, compared to age group (15-24), age group (25-34) has `r round(exp(2.26), 3)` times the odds of reporting good with significance in p value less than 0.05, whereas age group (35+) is not significantly different from the group (15-24) with a p value larger than 0.05 even though it has `r round(exp(1.98), 3)` the odds
of reporting good compared to group (15-24). 

It can seen that the coefficients in general linear mixed effects model differ from GEE model in the magnitude but not the significance. In GLLM model, `baseline`, `intervention`,
`time`, `age groups` leads to greater change in the odds of reporting good compared to GEE model while maintaining the equivalent sign of all coefficients. On the other hand, the Z statistics of all coefficients in GLLM model are less than (in magnitude) the Z statistics of the corresponding coefficients in GEE model, meaning the estimated coefficients are of less confidence though the reported significance stay the same at 0.05 level. 
